---
title: "Demo"
author:
- name: Jiayi Wang*
  affiliation:
  - &IMLS Institute for Molecular Life Sciences, University of Zurich, Switzerland
  - &SIB SIB Swiss Institute of Bioinformatics, University of Zurich, Switzerland
- name: Helena L Crowell*
  affiliation:
  - *IMLS
  - *SIB
- name: Mark D Robinson
  affiliation:
  - *IMLS
  - *SIB
output: 
  html_document:
    theme: simplex
    toc: true
    toc_depth: 2
    toc_float: true
date: "`r format(Sys.Date(), '%B %d, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE)
```

# preamble

## dependencies

```{r load-libs, message=FALSE, warning=FALSE}
library(scran)
library(scater)
library(igraph)
library(UpSetR)
library(ggplot2)
library(patchwork)
```

## simulation

In this demo, we will use synthetic data comprising 400 cells per sample (3 per 
2 conditions), 3 clusters, and 2,000 genes. We here source a script to simulate 
this scenario, but any data with analogous design could serve as input instead.

```{r sim}
# simulation
wcs <- list(t=80, s=80, b=0)
sim_rds <- tempfile(fileext=".rds")
args <- list(res=sim_rds)
source("code/00-get_sim.R")
(sim <- readRDS(args$res))
```

## processing

Next, we perform minimal filtering (excluding low-quality genes and cells) and
preprocessing steps (normalization, principal component analysis), as well as 
low- and high-resolution graph-based clustering. The resulting `cluster_lo/hi` 
(stored as `colData`) assignments, and biological variance component estimates
(`rowData` slot `bio`) are used by different feature scores/selections below.

```{r pro}
# keep genes with count > 1 in at least 10 cells,
# and cells with at least 10 detected genes
y <- counts(sim)
gs <- rowSums(y > 1) >= 10
cs <- colSums(y > 0) >= 10
sim <- sim[gs, cs]

# log-library size normalization, 
# feature selection, dimension reduction
sim <- logNormCounts(sim)
tbl <- modelGeneVar(sim, block=sim$sample_id)
hvg <- (rowData(sim)$bio <- tbl$bio ) > 0
sim <- runPCA(sim, subset_row=hvg, ncomponents=10)

# high- & low-resolution clustering
g <- buildSNNGraph(sim, use.dimred="PCA")
lo <- cluster_louvain(g, resolution=0)
hi <- cluster_louvain(g, resolution=2)
sim$cluster_lo <- factor(lo$membership)
sim$cluster_hi <- factor(hi$membership)
```

```{r plt-pcs, echo=FALSE, fig.width=8, fig.height=3}
plotPCA(sim, point_size=0.2, colour_by="group_id") +
      scale_color_manual("condition/group", values=c("royalblue", "tomato")) +
plotPCA(sim, point_size=0.2, colour_by="cluster_id") +
    scale_color_manual("cell type/cluster", values=c("cyan", "magenta", "gold")) +
plot_layout(nrow=1) & coord_equal() &
    guides(col=guide_legend(override.aes=list(alpha=1, size=2)))
```

# analysis

## scoring

We first source feature scoring scripts. Each defines a function that 
takes as input a `SingleCellExperiment`, and returns a score for each gene.

```{r get-sco}
# source scoring functions & assign to variables
sco <- list.files("code", "02-sco-", full.names=TRUE)
(ids <- gsub(".*-(.*)\\..*", "\\1", sco))
names(sco) <- names(sco_ids) <- sco_ids <- ids
for (. in sco_ids) { source(sco[.]); assign(., fun) }
# view exemplary function definition
HVG
```

Looping through these, we can compute a set of gene scores; here, using 
all strategies. In reality, only one or two scores would be sufficient.

```{r run-sco}
# run scoring
sco_res <- lapply(sco_ids, \(.) get(.)(sim))
# consolidate into table
names(sco_res)[2] <- ""
sco_res <- data.frame(sco_res)
# view results for a few genes
head(sco_res)
```

## selection

Analogous to above, we next source feature selection scripts. 
Each of these functions relies the above scores to select 2,000 genes. 
Specifically, selections with a `_` are based on the rank difference 
of two scores; remaining selections rely on one score only. 

```{r get-sel}
# source scoring functions & assign to variables
sel <- list.files("code", "03-sel_dat-", full.names=TRUE)
(ids <- gsub(".*-(.*)\\..*", "\\1", sel))
names(sel) <- names(sel_ids) <- sel_ids <- ids
for (. in sel_ids) { source(sel[.]); assign(., fun) }
# view exemplary function definition
tF_sPBDS
```

Again, we loop through all selection strategies to obtain a set of genes of each.
These functions default to selecting the top-2,000 ranked genes, which might be
reasonable in experimental data. However, because this is synthetic data and we
only simulated 2,000 genes, we subset selections to the top-200 ranked genes.

```{r run-sel}
# selection scripts expects a list of
# - gene_id = feature identifiers
# - sco_val = score values
sco_ret <- lapply(
    asplit(sco_res, 2), \(.) 
    list(gene_id=names(.), sco_val=.))
# run selections
sel_res <- lapply(sel_ids, \(.) get(.)(sco_ret)[seq_len(200)])
# view first couple features
sapply(sel_res, head)
```

## visuals

We here refrain from any quantitative evaluation and instead, briefly,
qualitatively/visually compare different feature selection strategies.

### agreement

The arguably simplest comparison is to compute and visualize, for all 
sets of strategies, the number of features unique/common between them.
E.g., we can observe that *random* selection results in the least overlaps,
and selections involvins *tF* and *sPVE*, respectively, share a few genes.

```{r plt-set, fig.width=6, fig.height=4}
# count how many genes are in common 
# between different selection strategies
upset(fromList(sel_res), sets=sel_ids, nintersects=20)
```

### reduction

Lastly, we can use feature subsets for principal component analysis (PCA) that, 
in practice, may underlie downstream analyses such as integration and clustering.
We also visually compare the top-2 PCs colored by group/experiment condition
and cluster/cell type, respectively. Ideally, we would like to observe that 
clusters are well separated but not groups. E.g., we can observe that HVGs lead
do to some extend separate clusters but also groups, while type- and type-state-
based selection result in a better separation of clusters and well-mixed groups.

```{r run-pca}
# do principal component analysis
# using different feature selections
sim_ <- lapply(sel_res, \(.) runPCA(sim, subset_row=.))
```

```{r plt-qcs, fig.width=12, fig.height=4}
# plotting
ps <- lapply(names(sim_), \(.) {
    p <- plotPCA(sim_[[.]], 
        # coloring by condition/group
        point_size=0.2, colour_by="group_id") +
        scale_color_manual(
            "condition/group", 
            values=c("royalblue", "tomato"))
    q <- plotPCA(sim_[[.]], 
        # coloring by cell type/cluster
        point_size=0.2, colour_by="cluster_id") +
        scale_color_manual(
            "cell type/cluster", 
            values=c("cyan", "magenta", "gold"))
    list(p+ggtitle(.), q)
}) |> Reduce(f=c)
# aesthetics
wrap_plots(ps) +
    plot_layout(nrow=2, byrow=FALSE, guides="collect") &
    guides(col=guide_legend(override.aes=list(alpha=1, size=2))) &
    coord_equal() & theme(
        legend.key.size=unit(0, "pt"),
        plot.title=element_text(hjust=0.5))
```

# appendix

::: {.callout-note icon=false, collapse=true}

## session

```{r sess-info}
#| code-fold: false
sessionInfo()
```

:::