---
title: "Compare neighbourhoods and clusters"

format:
  html:
    toc: true
    self-contained: true
    code-fold: true
    code-tools: true
    toc_float: true
    
author: Jiayi Wang*, Helena L. Crowell* and Mark Robinson
---

## Dependencies
```{r package}
suppressPackageStartupMessages({
    library(ggplot2)
    library(dplyr)
    library(miloDE)
    library(miloR)
    library(tidyr)
    library(ggh4x)
    library(patchwork) 
    library(tidytext) 
    library(stringr)
    library(ComplexHeatmap)
    library(gridExtra)
    library(data.table)
})
```


```{r setup, include=FALSE}
set.seed(123)
knitr::opts_knit$set(root.dir = "~/projects/type-state")
```

This vignette evaluates how muscat’s non-overlapping clusters correspond to miloDE’s overlapping neighborhoods. We do this by calculating, for each neighborhood, the proportion of cells assigned to each muscat cluster, using the Kang dataset as a case study. The subpopulations analyzed were derived using highly variable genes (HVG) and type-not-state feature selection (tF_sPBDS).

## Data loading

```{r data}
hvg <- readRDS("data/dat/02-rep/Kang18,HVG.rds")
tns <- readRDS("data/dat/02-rep/Kang18,tF_sPBDS.rds")
```

## miloDE vs. cluster_re 

```{r}
# get clusters/neighborhoods that 
# underwent differential testing
df_ds <- list(
    HVG=list(
        muscat="outs/dat/das-Kang18,HVG,DS_muscat.rds",
        miloDE="outs/dat/das-Kang18,HVG,DS_miloDE.rds"),
    tF_sPBDS=list(
        muscat="outs/dat/das-Kang18,tF_sPBDS,DS_muscat.rds",
        miloDE="outs/dat/das-Kang18,tF_sPBDS,DS_miloDE.rds"))
nh_id <- lapply(df_ds, lapply, \(.) {
    . <- unique(readRDS(.)$i_nhood)
    print(length(.)); return(.) })
```

```{r}
.prop <- \(x, sel) {
    # assign cells to neighborhoods
    #x <- hvg; sel <- "HVG"
    y <- assign_neighbourhoods(x, reducedDim_name="PCA")
    nh <- as.data.table(as.matrix(nhoods(y)))
    cols <- names(nh)
    nh$cluster_re <- x$cluster_re
    nh$cluster_id <- x$cluster_id
    nh$cell_id <- seq_len(nrow(nh))
    # quantify overlap between clusters & neighborhoods
    # (relative abundance of clusters in each neighborhood)
    nh <- melt(nh, 
        measure.vars = cols,
        id.vars = c("cluster_re", "cluster_id"), 
        variable.name = "neighborhood", value.name = "in_nn")
    ns <- seq_len(nlevels(unique(nh$neighborhood)))
    nh$neighborhood <- factor(nh$neighborhood, labels=ns)
    # filter for cells assigned to groups
    # that underwent differential testing
    i <- nh$cluster_re %in% nh_id[[sel]]$muscat
    j <- nh$neighborhood %in% nh_id[[sel]]$miloDE
    nh <- nh[(i & j) & nh$in_nn == 1, ] 
    nh$cluster_re <- droplevels(nh$cluster_re)
    nh$neighborhood <- droplevels(nh$neighborhood)
    nh[, size := .N, by = neighborhood]
    re <- nh[, .N, by = .(neighborhood, cluster_re, size)]
    id <- nh[, .N, by = .(neighborhood, cluster_id, size)]
    re[, proportion := N / size]
    id[, proportion := N / size]
    # re[, neighborhood := as.factor(neighborhood)]
    # id[, neighborhood := as.factor(neighborhood)]
    setnames(re, "cluster_re", "cluster")
    setnames(id, "cluster_id", "cluster")
    re[,type:="cluster_re"]
    id[,type:="cluster_id"]
    dt <- rbind(re,id)
    dt[, sel := sel]
    dt
}
 dt1 <- .prop(hvg,"HVG")
 dt2 <- .prop(tns,"tF_sPBDS")
 dt <- rbind(dt1,dt2)
 td <- dt[type=="cluster_re"]
 head(td)
```

```{r, fig.width=14, fig.height=4}
gg <- by(td, td$sel, function(dt) {
    #dt <- td[td$sel == "HVG", ]
    id <- dt$sel[1]
    print(length(unique(dt$cluster)))
    print(length(unique(dt$neighborhood)))
    dt <- select(dt, neighborhood, cluster, proportion)
    mx <- pivot_wider(dt, 
        names_from=cluster, 
        values_from=proportion, 
        values_fill=0)
    my <- `rownames<-`(as.matrix(mx[, -1]), mx[[1]])
    n <- "euclidean"; m <- "complete"
    xo <- colnames(my)[hclust(dist(t(my), n), m)$order]
    yo <- rownames(my)[hclust(dist(my, n))$order]
    # concordance
    p <- ggplot(dt, aes(cluster, neighborhood, fill=proportion)) + geom_tile() +
        scale_fill_gradient2(limits=c(0, 1), n.breaks=2, low="ivory", high="navy") +
        coord_equal(20/length(yo), expand=FALSE) +
        scale_y_discrete(limits=yo) +
        ggtitle(id) + theme_bw(6) + theme(
            axis.text=element_blank(),
            panel.grid=element_blank(),
            axis.ticks=element_blank(),
            plot.title=element_text(hjust=0.5),
            legend.key.width=unit(0.2, "lines"),
            legend.key.height=unit(0.4, "lines")) 
    # maxima
    nn <- length(unique(dt$neighborhood))
    df <- data.frame(cluster=names(val <- 1-apply(my, 2, max)), val)
    q <- ggplot(df, aes(cluster, 0, fill=val)) + 
        geom_tile(col="white", linewidth=0.1) + 
        theme_void() + theme(legend.position="none") + 
        scale_fill_gradient2(low="ivory", high="firebrick")
    # arrangement
    wrap_plots(p, q, ncol=1, heights=c(nn, 1)) & scale_x_discrete(limits=xo)
}) |> wrap_plots(nrow=1) + plot_layout(guides="collect") 
ggsave("Kang_cm.pdf", gg, width=8/2.54, height=6/2.54)
```

## miloDE vs. cluster_id
```{r, warning=FALSE, fig.width=12, fig.height=4}
tt <-  dt[dt$type == "cluster_id", ]
tt[, cluster:= factor(cluster, levels=unique(cluster))]
gg <- by(tt, tt$sel, function(dt) {
    id <- dt$sel[1]
    dt <- select(dt, neighborhood, cluster, proportion)
    mx <- pivot_wider(dt, names_from=cluster, 
        values_from=proportion, values_fill=0)
    my <- `rownames<-`(as.matrix(mx[, -1]), mx[[1]])
    xo <- colnames(my)[hclust(dist(t(my)))$order]
    yo <- rownames(my)[hclust(dist(my))$order]
    ggplot(dt, aes(cluster, neighborhood, fill=proportion)) + geom_tile(col=NA) +
        scale_fill_gradient2("prop.", limits=c(0, 1), n.breaks=2, low="ivory", high="navy") +
        scale_x_discrete(limits=paste(xo)) +
        scale_y_discrete(limits=paste(yo)) +
        coord_equal(expand=FALSE) +
        ggtitle(id) + 
        theme_bw(6) + theme(
            aspect.ratio=1,
            panel.grid=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks.y=element_blank(),
            legend.key.width=unit(0.2, "lines"),
            legend.key.height=unit(0.4, "lines"),
            axis.text.x=element_text(angle=45, hjust=1))
}) |> wrap_plots(nrow=1) + 
    plot_layout(guides="collect") 
ggsave("Kang_cm2.pdf", gg, width=7/2.54, height=5/2.54)
```



