---
title: "Investigating the Relationship Between Harmony Integration and Feature Selection Methods"

format:
  html:
    toc: true
    self-contained: true
    code-fold: true
    code-tools: true
    toc_float: true
    
author: 
  Jiayi Wang$^*$, Helena L. Crowell$^*$, and Mark D. Robinson 
  Institute for Molecular Life Sciences, University of Zurich, Switzerland;  
  SIB Swiss Institute of Bioinformatics, University of Zurich, Switzerland  
  
---



## Dependencies
```{r}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(harmony)
    library(igraph)
    library(patchwork)
    library(ggplot2)
    library(scater)
    library(scran)
    library(tidyr)
    library(lisi)
    library(knitr)
    library(DT)
    library(data.table)
    library(rstudioapi)
})

```

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_knit$set(root.dir = "/Users/jiayi/type-state")
```

```{r function}

# plot
.plot <- \(sce, sel) {
    p1 <- plotPCA(sce, color_by = "cluster_id") + ggtitle(sel) +
    plotPCA(sce, color_by = "group_id") + 
    scale_color_brewer(palette = "Set1") +
    labs(col = "group_id")

    p2 <- plotReducedDim(sce, dimred = "HARMONY", color_by = "cluster_id") +
        ggtitle(paste0(sel, " + Harmony")) + 
        plotReducedDim(sce, dimred = "HARMONY", color_by = "group_id") + 
        scale_color_brewer(palette = "Set1")  + labs(col = "group_id")
    
    p1 / p2
}

# LISI 
.lisi <- \(x, rd){
    y <- reducedDim(x, rd)
    cd <- colData(x)
    by <- c(
        LISI_g="group_id", 
        LISI_k="cluster_id")
    res <- sapply(by, \(.) {
        res <- compute_lisi(y, cd, .)
        n <- length(unique(cd[[.]]))
        (mean(res[[1]])-1)/(n-1)
    })
    data.frame(
        row.names=NULL, 
        sta=names(res), 
        sta_val=res)
}

.compute_lisi <- \(sce, sel) {
    lisi1 <- .lisi(sce, "PCA")
    lisi2 <- .lisi(sce, "HARMONY")
    lisi1$method <- sel
    lisi2$method <- paste0(sel, "+Harmony")
    tbl <- rbind(lisi1, lisi2)
    colnames(tbl) <- c("metric", "value", "method")
    tbl %>% 
        pivot_wider(names_from = metric, values_from = value)
}
```

## Simulated data

We selected one simulated dataset (t = 1, s = 1) to illustrate the relationship between type-not-state feature selection and data integration. For this example, we used Harmony integration. First, we compared dimensionality reduction results (PCA) based on different feature selection strategies, both before and after integration.

We then used the `LISI` score (as implemented in Harmony) to quantify the mixing of cell types or clusters (`LISI_k`) and cell states or groups (`LISI_g`). A higher `LISI_g` score indicates better mixing of cell states, while a lower `LISI_k` score reflects better separation of clusters.


### HVG vs. HVG + Harmony

```{r sim_hvg, fig.width = 10, fig.height = 10, warning=FALSE, message=FALSE}
# HVG 
sce <- readRDS("data/sim/02-rep/t100,s100,b0,HVG.rds")
sel <- "HVG"
# HVG + harmony
sce <- RunHarmony(sce,
  group.by.vars = "sample_id",
  reduction.use = "PCA",
  verbose = FALSE)

.plot(sce,sel)
df1 <- .compute_lisi(sce, sel)
df1
```

### Random vs. Random + Harmony

```{r sim_random, fig.width = 10, fig.height = 10, warning=FALSE, message=FALSE}
# random 
sce <- readRDS("data/sim/02-rep/t100,s100,b0,random.rds")
sel <- "Random"
# random + harmony
sce <- RunHarmony(sce,
  group.by.vars = "sample_id",
  reduction.use = "PCA",
  verbose = FALSE)

.plot(sce,sel)
df2 <- .compute_lisi(sce, sel)
df2
```


### tF vs. tF + Harmony
```{r  sim_tF, fig.width = 10, fig.height = 10, warning=FALSE, message=FALSE}
sce <- readRDS("data/sim/02-rep/t100,s100,b0,tF.rds")
sel <- "tF"
# random + harmony
sce <- RunHarmony(sce,
  group.by.vars = "sample_id",
  reduction.use = "PCA",
  verbose = FALSE)

.plot(sce,sel)
df3 <- .compute_lisi(sce, sel)
df3
```


### tF_sPBDS vs. tF_sPBDS + Harmony

```{r sim_tF_sPBDS, fig.width = 10, fig.height = 10, message=FALSE}
sce <- readRDS("data/sim/02-rep/t100,s100,b0,tF_sPBDS.rds")
sel <- "tF_sPBDS"
# random + harmony
sce <- RunHarmony(sce,
  group.by.vars = "sample_id",
  reduction.use = "PCA",
  verbose = FALSE)

.plot(sce,sel)
df4 <- .compute_lisi(sce, sel)
df4
```


### LISI score overall comparison

Here we summarized the LISI scores for all feature selection methods, with and without Harmony integration. Among all tested methods, tF_sPBDS+Harmony consistently achieved the best trade-off, with the highest LISI_g (0.87) and low LISI_k (0.04). Notably, tF_sPBDS alone (without Harmony) already substantially improved group mixing, outperforming random+Harmony and all other feature selections without integration. However, its type separation was suboptimal without integration, suggesting that integration remains necessary for optimal performance. 


```{r}
df <- do.call(rbind, list(df1,df2,df3,df4))
setDT(df)
datatable(df, options = list(pageLength = 10), rownames = FALSE)
```


```{r fig.height=6, fig.width=4.5}
df[, feature_selection := gsub("\\+Harmony", "", method)]
df[, Harmony := ifelse(grepl("Harmony", method), "Yes", "No")]
# pal <- c(
#   "HVG" = "#FDB863",           # light orange
#   "HVG+Harmony" = "#E66101",  # dark orange
#   "Random" = "#B2ABD2",        # light lavender
#   "Random+Harmony" = "#5E3C99",# dark purple
#   "tF" = "#A6DBA0",            # light green
#   "tF+Harmony" = "#1B7837",   # dark green
#   "tF_sPBDS" = "#c6dbef",      # pale blue
#   "tF_sPBDS+Harmony" = "#3182bd" # darker blue
# )
# df$method <- factor(df$method, levels=df$method)
# names(pal) <- df$method
# ggplot(df, aes(x = reorder(method, -LISI_g), y=LISI_g, fill = method)) +
#   geom_bar(stat = "identity") +
#   scale_fill_manual(values = pal) +
#   theme_bw() +
#   labs(x = NULL, y = "LISI_g", fill = "Method") +
#   theme(axis.text.x = element_blank(),
#         axis.ticks = element_blank()) +
#   ggtitle("Mixing of cell states/groups")

pal <- c(
  "#F4894B", 
  "#FAED6D",  
  "#D35FB7", 
  "#54C4DA"   
)

p1 <- ggplot(df, aes(x = reorder(feature_selection, LISI_g), y = LISI_g, 
               fill = feature_selection, alpha = Harmony)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  scale_fill_manual(values = pal) +
  theme_bw() +
  labs(x = NULL, y = "LISI_g", fill = "Method") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Mixing of cell states/conditions") +
  scale_alpha_manual(values = c("No" = 0.6, "Yes" = 1)) 


p2 <- ggplot(df, aes(x = reorder(feature_selection, -LISI_k), y = LISI_k, 
               fill = feature_selection, alpha = Harmony)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  scale_fill_manual(values = pal) +
  theme_bw() +
  labs(x = NULL, y = "LISI_k", fill = "Method") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Mixing of cell types/clusters") +
  scale_alpha_manual(values = c("No" = 0.6, "Yes" = 1)) 


p1 / p2 + plot_layout(guides="collect")
```


## Kang's datatset

We performed a similar comparison using Kang’s dataset, but visualized the results with UMAPs.

### HVG vs. HVG + Harmony

```{r kang_hvg, fig.width = 10, fig.height = 8, warning=FALSE, message=FALSE}
## HVG
sce <- readRDS("data/dat/01-pro/Kang18.rds")
.f <- \(sce, sel) {
    p1 <- plotUMAP(sce, color_by = "cluster_id") + ggtitle(sel) +
    plotUMAP(sce, color_by = "group_id") + 
        scale_color_brewer(palette = "Set1") +
        labs(col = "group_id")


    ## HVG + Harmony
    sce <- RunHarmony(sce,
      group.by.vars = "sample_id",
      reduction.use = "PCA",
      verbose = FALSE)
    sce <- runUMAP(sce, dimred = "HARMONY")
    
    
    p2 <- plotUMAP(sce, color_by = "cluster_id") + ggtitle(paste0(sel, " + Harmony")) +
        plotUMAP(sce, color_by = "group_id") +
            scale_color_brewer(palette = "Set1") +
            labs(col = "group_id")
    
    p <- p1 / p2
    
    ## LISI score
    lisi1 <- .lisi(sce, "PCA")
    lisi2 <- .lisi(sce, "HARMONY")
    lisi1$method <- sel
    lisi2$method <- paste0(sel,"+Harmony")
    
    
    tbl <- rbind(lisi1, lisi2)
    colnames(tbl) <- c("metric", "value", "method")
    df <- tbl %>% 
      pivot_wider(names_from = metric, values_from = value)
    return(list(p=p, df=df))
}
res <- .f(sce, "HVG")
res$p
df1 <- res$df
df1
```

### Random vs. random + Harmony
```{r kang_random, fig.width = 10, fig.height = 8, warning=FALSE, message=FALSE}
sce <- readRDS("data/dat/02-rep/Kang18,random.rds")
res <- .f(sce, "Random")
res$p
df2 <- res$df
df2
```

### tF vs. tF + Harmony

```{r kang_tf, fig.width = 10, fig.height = 8, warning=FALSE, message=FALSE}
sce <- readRDS("data/dat/02-rep/Kang18,tF.rds")
res <- .f(sce, "tF")
res$p
df3 <- res$df
df3
```


### tF_sPBDS vs. tF_sPBDS + Harmony 

```{r kang_tF_sPBDS, fig.width = 10, fig.height = 8, message=FALSE}
## tF_sPBDS
sce <- readRDS("data/dat/02-rep/Kang18,tF_sPBDS.rds")
res <- .f(sce, "tF_sPBDS")
res$p
df4 <- res$df
df4
```

### LISI score overall comparison

Here, the performance gains with integration were even more pronounced. `tF_sPBDS+Harmony` outperformed standard `HVG+Harmony` and `tF_sPBDS`, achieving higher group mixing (`LISI_g` = 0.64) with comparable type separation (`LISI_k` = 0.028). We noticed that the performance gain was even larger than in the simulated dataset. This is likely because our simulated data did not explicitly include technical or sample-specific effects, whereas real dataset probably contains technical variation that Harmony is designed to correct. 

In conclusion, we proposed that type-not-state feature selection is complementary to integration. By selecting biologically relevant features, `tF_sPBDS` provides a better input embedding for integration.

```{r}
df <- do.call(rbind, list(df1,df2,df3,df4))
setDT(df)
datatable(df, options = list(pageLength = 10), rownames = FALSE)
```

```{r fig.height=6, fig.width=4.5}
df[, feature_selection := gsub("\\+Harmony", "", method)]
df[, Harmony := ifelse(grepl("Harmony", method), "Yes", "No")]
pal <- c(
  "#F4894B", 
  "#FAED6D",  
  "#D35FB7", 
  "#54C4DA"   
)

p1 <- ggplot(df, aes(x = reorder(feature_selection, LISI_g), y = LISI_g, 
               fill = feature_selection, alpha = Harmony)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  scale_fill_manual(values = pal) +
  theme_bw() +
  labs(x = NULL, y = "LISI_g", fill = "Method") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Mixing of cell states/groups") +
  scale_alpha_manual(values = c("No" = 0.6, "Yes" = 1)) 


p2 <- ggplot(df, aes(x = reorder(feature_selection, -LISI_k), y = LISI_k, 
               fill = feature_selection, alpha = Harmony)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  scale_fill_manual(values = pal) +
  theme_bw() +
  labs(x = NULL, y = "LISI_k", fill = "Method") +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank()) +
  ggtitle("Mixing of cell types/clusters") +
  scale_alpha_manual(values = c("No" = 0.6, "Yes" = 1)) 


p1 / p2 + plot_layout(guides="collect")
```


## Session Info
```{r}
sessionInfo()
```

